---
title: "Leak Detection N-Grams"
author: "Kat Bardash"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message=FALSE)
library(tidytext)
library(ggplot2)
```

# Purpose
- The purpose of this file is conduct text analysis on the leak_detection_field_technology column in the foundational_06a_loss_and_leak_detection file. Distributions of time between testing and age to replace small meters in years.
- This analysis creates token, bi-grams, tri-grams, and four-grams of this text field.
- The data can be found http://cowaterefficiency.com/unauthenticated_home with permission. Once in the portal, all report years (2013-2017) were selected as well as all water providers.

## Recommendations
- Water detection technology could be a useful factor to correlate and predict with water loss. The current input method does not easily allow for this type of analysis.
- Our suggestion is to create a drop-down menu for this particular part of the reporting with the applicable leak detection technologies. 

```{r}
# read in data
found_6a_lossleak <- read.csv("EffDataPortal_Output_User690_20181112192716/foundational/foundational_06a_loss_and_leak_det.csv", stringsAsFactors = FALSE)
hist(found_6a_lossleak$time_between_tests_largemeters_years, main="Distribution of Years Between Tests For Large Meters")
hist(found_6a_lossleak$replace_age_small_meters_years, main="Distribution of Years Between Tests For Small Meters")
```

###  Text Analysis of Leak Detection Type with N-grams
```{r}

leaks <- found_6a_lossleak
head(leaks[order(leaks$ce_annual_ndx),], n=3)

leaks_short <- leaks[,c(1,9 )]
head(leaks_short, n=3)
library(dplyr)
library(tidytext)
leaks_short2<- leaks_short %>% unnest_tokens(word, leak_detection_field_technology)

# get rid of stop words
tidy_leaks <- leaks_short2 %>% anti_join(stop_words)


```

### Most used Words
- These are the top 20 most common words typed into the leak detection technology field. 
```{r}

head(tidy_leaks %>% count(word, sort=TRUE), n=20)

```

###  Finding n-grams
- These are the top 10 bi, tri, and 4-grams for the leak detection technology column.
```{r}

#2-gram
tidy_bigram <- leaks_short %>% unnest_tokens(bigram,leak_detection_field_technology, token="ngrams", n=2)
bill10 <- head(tidy_bigram %>% count(bigram, sort=TRUE), n=10)
# remove NA ( reorder)
bill10 <- bill10[!is.na(bill10$bigram),]
bill10 <- bill10 %>% mutate(bigram=reorder(bigram, n))

   ggplot(bill10, aes(bigram, n))+
  geom_col()+
  coord_flip()+
  ggtitle("Top 8 Bigrams From Leak Detection Field Technology")
```
```{r}
#3-gram
tidy_trigram <- leaks_short %>% unnest_tokens(trigram,leak_detection_field_technology, token="ngrams", n=3)
tri10 <- head(tidy_trigram %>% count(trigram, sort=TRUE), n=10)

# remove NA ( reorder)
tri10 <- tri10[!is.na(tri10$trigram),]
tri10 <- tri10 %>% mutate(trigram=reorder(trigram, n))

   ggplot(tri10, aes(trigram, n))+
  geom_col()+
  coord_flip()+
  ggtitle("Top 9 Trigrams From Leak Detection Field Technology")
```
#4-gram

```{r}
tidy_fourgram <- leaks_short %>% unnest_tokens(fourgram,leak_detection_field_technology, token="ngrams", n=4)
four10 <- head(tidy_fourgram %>% count(fourgram, sort=TRUE), n=10)

# remove NA ( reorder)
four10 <- four10[!is.na(four10$fourgram),]
four10 <- four10 %>% mutate(fourgram=reorder(fourgram, n))

   ggplot(four10, aes(fourgram, n))+
  geom_col()+
  coord_flip()+
  ggtitle("Top 9 Fourgrams From Leak Detection Field Technology")
```
