---
title: "waterByCity"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(sqldf)
library(data.table)
library(dplyr)
library(tidyr)
library(ggplot2)
library(rlist)
library(lubridate)
```

# Purpose
- The purpose here is to understand water production by city and population served using HB-1051 data. Data was obtained from [CWCB Water Efficiency Data Portal](http://cowaterefficiency.com/unauthenticated_home) 
All years and water providers were retrieved

## Read in data
- Files used:
    + water_use_02_production.csv ([HB 1051 Data](http://cowaterefficiency.com/unauthenticated_home))
    + overview.csv ([HB 1051 Data](http://cowaterefficiency.com/unauthenticated_home))
    + normalizing_03_population.csv ([HB 1051 Data](http://cowaterefficiency.com/unauthenticated_home))
    + historical-census-data-counties.csv ([Historical Census Data](https://demography.dola.colorado.gov/population/data/historical_census/))
    + historical-census-data-municipalities.csv ([Historical Census Data](https://demography.dola.colorado.gov/population/data/historical_census/))
```{r read data}
production_filepath = "/Users/dsciacca/Documents/Data Analytics Challenge/Water Efficiency Data - CWCB/water_use/water_use_02_production.csv"
overview_filepath = "/Users/dsciacca/Documents/Data Analytics Challenge/Water Efficiency Data - CWCB/overview/overview.csv"
pop_filepath = "/Users/dsciacca/Documents/Data Analytics Challenge/Water Efficiency Data - CWCB/normalizing/normalizing_03_population.csv"
water_production = read.csv(production_filepath, header = TRUE, stringsAsFactors = FALSE)
water_overview = read.csv(overview_filepath, header = TRUE, stringsAsFactors = FALSE)
water_pop = read.csv(pop_filepath, header = TRUE, stringsAsFactors = FALSE)
```

## Clean up water_production file
### Convert all units to million gallons
```{r clean units}
#first place all dif measurments in separate dfs
water_prod_acre = sqldf("SELECT * FROM water_production WHERE units = 'Acre-feet'")
water_prod_ccf = sqldf("SELECT * FROM water_production WHERE units = 'CCF (100 cubic feet)'")
water_prod_milgal = sqldf("SELECT * FROM water_production WHERE units = 'Gallons, Millions'")
water_prod_thougal = sqldf("SELECT * FROM water_production WHERE units = 'Gallons, Thousands'")
water_prod_gal = sqldf("SELECT * FROM water_production WHERE units = 'Gallons'")

#1 acre-foot = 325851 gal = 0.325851 mil gal
water_prod_acre$produced = water_prod_acre$produced * 0.325851
water_prod_acre$exported = water_prod_acre$exported * 0.325851
water_prod_acre$water_in_system = water_prod_acre$water_in_system * 0.325851

#1 ccf = 748.052 gal = 0.000748052 mil gal
water_prod_ccf$produced = water_prod_ccf$produced * 0.000748052
water_prod_ccf$exported = water_prod_ccf$exported * 0.000748052
water_prod_ccf$water_in_system = water_prod_ccf$water_in_system * 0.000748052

#thousand gal to mil gal
water_prod_thougal$produced = water_prod_thougal$produced * 0.001
water_prod_thougal$exported = water_prod_thougal$exported * 0.001
water_prod_thougal$water_in_system = water_prod_thougal$water_in_system * 0.001

#gal to mil gal
water_prod_gal$produced = water_prod_gal$produced / 1000000
water_prod_gal$exported = water_prod_gal$exported / 1000000
water_prod_gal$water_in_system = water_prod_gal$water_in_system / 1000000

#recombine datasets - effectively drops unspecified and blank unit rows, 8 total
water_prod = rbind(water_prod_gal, water_prod_milgal, water_prod_thougal, water_prod_acre, water_prod_ccf)

#drop units column
water_prod = water_prod[-c(7)] 
```

## Merge datasets 
### First merge water production with overview on ce_annual_ndx to get city, county, and ce_index, etc. columns. Then, merge with population dataset, again on ce_annual_ndx to get population data
```{r merge files}
#merge production and overview
water_prod_city = merge(water_prod, water_overview, all.x = TRUE, by.x = "ce_annual_ndx", by.y = "ce_annual_ndx")

#merge with population dataset
water_prod_city_pop = merge(water_prod_city, water_pop, all.x = TRUE, by.x = "ce_annual_ndx", by.y = "ce_annual_ndx")
```
##Clean data
### Drop extra columns, then remove rows with NA in produced column
```{r drop columns}
# Drop columns (remarks, pop sources, provider comments, transient pop description)
water_prod_city_pop = water_prod_city_pop[, -c(7, 15, 20, 22, 23, 24)]

water_prod_city_pop = as.data.table(water_prod_city_pop)
water_prod_city_pop = na.omit(water_prod_city_pop, cols = "produced")
water_prod_city_pop = as.data.frame(water_prod_city_pop)
```


## Separate datasets into different water types and remove duplicate rows
```{r seperate to water types}
water_prod_city_pop_potable = sqldf("SELECT * FROM water_prod_city_pop WHERE water_type_index = 1")
water_prod_city_pop_npraw = sqldf("SELECT * FROM water_prod_city_pop WHERE water_type_index = 2")
water_prod_city_pop_npreuse = sqldf("SELECT * FROM water_prod_city_pop WHERE water_type_index = 3")

#remove duplicate rows - all but npreuse had extra rows, including just to be sure
water_prod_city_pop_potable = unique(water_prod_city_pop_potable)
water_prod_city_pop_npraw = unique(water_prod_city_pop_npraw)
water_prod_city_pop_npreuse = unique(water_prod_city_pop_npreuse)
```

## Detect outliers to make visualization more useful
```{r detect and remove outliers}
#first find outlier values
potable_outliers = boxplot.stats(water_prod_city_pop_potable$produced)$out
npraw_outliers = boxplot.stats(water_prod_city_pop_npraw$produced)$out
npreuse_outliers = boxplot.stats(water_prod_city_pop_npreuse$produced)$out

#obtain indeces of outliers in each dataset
potable_outlier_ndx = match(potable_outliers, water_prod_city_pop_potable$produced)
npraw_outlier_ndx = match(npraw_outliers, water_prod_city_pop_npraw$produced)
npreuse_outlier_ndx = match(npreuse_outliers, water_prod_city_pop_npreuse$produced)

#create boxplots with outliers removed
par(mfrow=c(1,3))
boxplot(water_prod_city_pop_potable$produced[-potable_outlier_ndx], main = "Potable water produced", ylab = "Gallons of Water Produced (millions)")
boxplot(water_prod_city_pop_npraw$produced[-npraw_outlier_ndx], main = "Non-potable Raw water produced", ylab = "Gallons of Water Produced (millions)")
boxplot(water_prod_city_pop_npreuse$produced[-npreuse_outlier_ndx], main = "Non-potable Reuse water produced", ylab = "Gallons of Water Produced (millions)")
```

## Plot water produced against population served
```{r plots}
plot(water_prod_city_pop_potable$pop_served[-potable_outlier_ndx], water_prod_city_pop_potable$produced[-potable_outlier_ndx], main = "Potable Water Produced vs. Population Served", xlab = "Population Served", ylab = "Gallons of Water Produced (millions)")
plot(water_prod_city_pop_npraw$pop_served[-npraw_outlier_ndx], water_prod_city_pop_npraw$produced[-npraw_outlier_ndx], main = "Non-potable Raw Water Produced vs. Population Served", xlab = "Population Served", ylab = "Gallons of Water Produced (millions)")
plot(water_prod_city_pop_npreuse$pop_served[-npreuse_outlier_ndx], water_prod_city_pop_npreuse$produced[-npreuse_outlier_ndx], main = "Non-potable Reuse Water Produced vs. Population Served", xlab = "Population Served", ylab = "Gallons of Water Produced (millions)")
```

## Build Water Production Model
```{r Water Production model}
#create new df without outliers
water_prod_city_pop_potable_noutlier = water_prod_city_pop_potable[-potable_outlier_ndx,]

#fit model on new df, use pop to predict water produced
fit = lm(produced ~ pop_served, data = water_prod_city_pop_potable_noutlier)

#check significance of the model
summary(fit)

#plot the model to examine adequacy
plot(fit)
```

## Build Population Model (Code Adapted from Kellen)
```{r Population model}
#Load data
county_dt <- fread("/Users/dsciacca/Documents/Data Analytics Challenge/historical-census-data-counties.csv")
municipality_dt <- fread("/Users/dsciacca/Documents/Data Analytics Challenge/historical-census-data-municipalities.csv")

# Remove data with 0 population
county_dt <- county_dt[county_dt$`Total Population` != 0, ]
municipality_dt <- municipality_dt[municipality_dt$`Total Population` != 0, ]

# Only keep citys with 5 or more population values
county_dt<-county_dt[,count:=.N,by=`AREA(S)`][(count>=5)]
municipality_dt<-municipality_dt[,count:=.N,by=`AREA(S)`][(count>=5)]

# Find the unique areas in the data
counties <- unique(county_dt$`AREA(S)`)
municipalities <- unique(municipality_dt$`AREA(S)`)

# Create an empty list
county_mods <- list()
municipality_mods <- list()

# Loop a 3rd order polynomial fit for each area and save results in the empty list
for (i in 1:length(counties)){
  county_mods[[i]] <- lm(`Total Population` ~ poly(YEAR, 3), data = county_dt[county_dt$`AREA(S)` == counties[i]])
}
for (i in 1:length(municipalities)){
  municipality_mods[[i]] <- lm(`Total Population` ~ poly(YEAR, 3), data = municipality_dt[municipality_dt$`AREA(S)` == municipalities[i]])
}

# Create an empty prediction list
county_preds = NULL
municipality_preds = NULL

# Future years to predict population
fy <- data.frame('YEAR' = c(2020, 2030, 2040, 2050))

# Add the unique areas to the prediction list and empty prediction years
county_preds<- data.table(counties)
county_preds$Y2020_pop = NA
county_preds$Y2030_pop = NA
county_preds$Y2040_pop = NA
county_preds$Y2050_pop = NA

municipality_preds<- data.table(municipalities)
municipality_preds$Y2020_pop = NA
municipality_preds$Y2030_pop = NA
municipality_preds$Y2040_pop = NA
municipality_preds$Y2050_pop = NA

# Loop predictions through and save results
for (i in 1:length(counties)){
  county_preds$Y2020_pop[i] <- predict(county_mods[[i]], fy)[1]
  county_preds$Y2030_pop[i] <- predict(county_mods[[i]], fy)[2]
  county_preds$Y2040_pop[i] <- predict(county_mods[[i]], fy)[3]
  county_preds$Y2050_pop[i] <- predict(county_mods[[i]], fy)[4]
}

for (i in 1:length(municipalities)){
  municipality_preds$Y2020_pop[i] <- predict(municipality_mods[[i]], fy)[1]
  municipality_preds$Y2030_pop[i] <- predict(municipality_mods[[i]], fy)[2]
  municipality_preds$Y2040_pop[i] <- predict(municipality_mods[[i]], fy)[3]
  municipality_preds$Y2050_pop[i] <- predict(municipality_mods[[i]], fy)[4]
}
```

## Predict Future Water Production
```{r Predict Water Production}
#add empty water prediction columns
county_preds$Y2020_water = NA
county_preds$Y2030_water = NA
county_preds$Y2040_water = NA
county_preds$Y2050_water = NA

municipality_preds$Y2020_water = NA
municipality_preds$Y2030_water = NA
municipality_preds$Y2040_water = NA
municipality_preds$Y2050_water = NA

#make water production predictions and save results
county_preds$Y2020_water = predict(fit, data.frame("pop_served" = county_preds$Y2020_pop))
county_preds$Y2030_water = predict(fit, data.frame("pop_served" = county_preds$Y2030_pop))
county_preds$Y2040_water = predict(fit, data.frame("pop_served" = county_preds$Y2040_pop))
county_preds$Y2050_water = predict(fit, data.frame("pop_served" = county_preds$Y2050_pop))

municipality_preds$Y2020_water = predict(fit, data.frame("pop_served" = municipality_preds$Y2020_pop))
municipality_preds$Y2030_water = predict(fit, data.frame("pop_served" = municipality_preds$Y2030_pop))
municipality_preds$Y2040_water = predict(fit, data.frame("pop_served" = municipality_preds$Y2040_pop))
municipality_preds$Y2050_water = predict(fit, data.frame("pop_served" = municipality_preds$Y2050_pop))

# Write our data to csv
write.csv(county_preds, "/Users/dsciacca/Documents/Data Analytics Challenge/water_population_predictions_counties.csv", row.names = FALSE)

write.csv(municipality_preds, "/Users/dsciacca/Documents/Data Analytics Challenge/water_population_predictions_municipalities.csv", row.names = FALSE)
```
